Type Flattening Obfuscation

### 2020-01-28 paper/tool.tex
shift left/right operators

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[a4paper]{llncs}
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}
\usepackage[english]{babel}
\usepackage[autostyle]{csquotes}
\usepackage[sorting=ynt,sortcites=true,backend=bibtex8,doi=false,maxnames=4,isbn=false,firstinits=true]{biblatex}
\addbibresource{reference}

@@

\begin{document}
%
\title{Type Flattening Obfuscation}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Ta Thanh Dinh}
%
\authorrunning{Ta Thanh Dinh}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
	\email{tathanhdinh@gmail.com}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
	Beside data and control flow, high-level types are also important in program analysis,
	indeed type recovery is an essential step in binary code decompilation.
	For the purpose of anti-decompilation, the paper introduces a novel obfuscation
	technique which makes types become harder to be reconstructed.

@@

	% We show that the techniques used for type reconstruction can be bypassed,
	% and we claim that the bypass is always possible given \emph{semantics gap} between
	% type systems of high-level and machine language.
	% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}
An important step in binary code \emph{decompilation}~\cite{cifuentes_reverse_1994} is to
detect high-level objects (e.g. functions, variables) from low-level machine dependent
objects~\cite{balakrishnan_divine_2007} (e.g. registers, raw memory accesses, machine
instructions) before annotate them with types of the supposed original language. Because
most of information about high-level types is lost in compilation, the binary code
\emph{type recovery} requires special techniques, a survey up until 2015 can be
referenced in~\cite{caballero_type_2016}.

In the opposite direction, binary code \emph{obfuscation} is to make the decompilation (or in
general code analysis) harder. Since the data and control flow are principal elements for the
program analysis, current obfuscation techniques~\cite{collberg_surreptitious_2009} focus mostly
on them. But to the best of our knowledge, there is still no explicit effort in obfuscating
high-level types. There would be several reasons for this lack of effort.

\emph{First}, type obfuscation is unsurprisingly a side effect of data or control flow obfuscation.
Indeed, type recovery algorithms need both data and control flow to build type constraints,
if any of them is hidden
then the algorithms cannot work correctly. Or if the function boundary is not found
(because of anti-disassembing tricks, for example), then high-level objects cannot be recognized.
\emph{Second}, high-level types seem too coarse to be worthy
of being protected, in many cases just knowing certain values of the input which make the program
exploitable is enough. But knowledge about types expands attack surfaces because more analysis
can be proceeded, beside decompilation see the survey~\cite{caballero_type_2016} for a
more complete list.
% We exploit the \emph{semantics gap}

\section{Type flattening compiler}
We present \emph{uCc}, an open source C compiler that explicitly obfuscates high-level types.
The compiler is implemented in Rust, its source code and a short user guide are given at (cite).
Our effort is to attack the core of type recovery algorithms,
these are the data flow used to build type constraints, and the \emph{semantic gap} between types in the high-level language and their
representations in the low-level machine code. Beside custom tricks, \emph{uCc} uses extensively
obfuscation transformations based on Mixed Boolean Arithmetic expressions~\cite{eyrolles_obfuscation_2017,zhou_information_2007}.
To make some code diversity, the compiler is \emph{probabilistic}: given a source
code, \emph{uCc} generates each time a different (but computationally equivalent) output machine code.

We test outputs of \emph{uCc} against some decompilers, table~\ref{description} describes briefly
tested functions and their types in source code, table~\ref{flattening} shows results of using
decompilers to decompile binaries generated by \emph{uCc}. The test suite is also available at
the repository.

\begin{table}
	\begin{center}
		\caption{Functions and original types}\label{description}
		\begin{tabular}{|l|l|}
			\hline
			Description & Function type \\
			\hline
			identity & \texttt{int id(int)} \\
			\hline
			division & \texttt{int div(short, char)} \\
			\hline
			modular & \texttt{int mod(int)} \\
			\hline
			increase pointer then dereference & \texttt{int inc\_deref(int*)} \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

\begin{table}
	\begin{adjustwidth}{-6.21in}{-5.95in}
		\begin{center}
			% \centering
			\caption{Type flattening effect}\label{flattening}
			% \begin{adjustbox}{max width=1.5\textwidth}
			\begin{tabular}{|l|l|l|l|l|}
				\hline
				Original type & Hex-Rays & JEB & Ghidra & Snowman \\
				\hline
				\texttt{int id(int)} & \texttt{int64 id(int64)} & \texttt{ulong id(ulong)} &
				\texttt{ulong id(void)} & \texttt{int64 id(int64)} \\
				\hline
				\texttt{short div(short, char)} & \texttt{int64 div(uint64, uint64)} & \texttt{div\_t div(int, int)} &
				\texttt{div\_t div(int, int)} & \texttt{int64\_t div(int64, int64)} \\
				\hline
				\texttt{char mod(short, char)} & \texttt{int64 mod(int64, int64)} & \texttt{ulong mod(ulong, ulong)} &
				\texttt{ulong mod(void)} & \texttt{int64 mod(int64, int64)} \\
				\hline
				\texttt{int inc\_deref(int*)} & \texttt{int64 inc\_deref(int64)} & \texttt{ulong inc\_deref(ulong)} &
				\texttt{ulong inc\_deref(undefined8)} & \texttt{int64 inc\_deref(uint64)} \\
				\hline
			\end{tabular}
		\end{center}
	\end{adjustwidth}
\end{table}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[a4paper]{llncs}
%
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}
\usepackage[english]{babel}
\usepackage[autostyle]{csquotes}
\usepackage[sorting=ynt,sortcites=true,backend=bibtex8,doi=false,maxnames=4,isbn=false,firstinits=true]{biblatex}
\addbibresource{reference}

@@

\begin{document}
%
\title{Type Flattening Obfuscation}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Ta Thanh Dinh}
%
\authorrunning{Ta Thanh Dinh}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
	\email{tathanhdinh@gmail.com}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
	Beside data and control flow, high-level types are also important in program analysis,
	indeed type recovery is an essential step in binary code decompilation.
	For the purpose of anti-decompilation, the paper presents a novel obfuscation
	technique which makes types become harder to be reconstructed.

@@

	% We show that the techniques used for type reconstruction can be bypassed,
	% and we claim that the bypass is always possible given \emph{semantics gap} between
	% type systems of high-level and machine language.
	% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}
An important step in binary code \emph{decompilation}~\cite{cifuentes_reverse_1994} is to
detect high-level objects (e.g. functions, variables) from low-level machine dependent
objects~\cite{balakrishnan_divine_2007} (e.g. registers, raw memory accesses, machine
instructions) before annotate them with types of the "assumed" original language.
In a statically typed language like C, the compiler does not preserve the high-level
type information in the generated machine code,
% Because most of information about high-level types is lost in compilation,
% the machine code
then the \emph{type recovery} (or type reconstruction) from machine code requires
special techniques, a survey up until 2015 can be referenced in~\cite{caballero_type_2016}.

In the opposite direction, binary code \emph{obfuscation} is to make the decompilation (or in
general code analysis) harder. Since the data and control flow are principal elements for the
program analysis, current obfuscation methods~\cite{collberg_surreptitious_2009} focus mostly
on them. But to the best of our knowledge, there is still no explicit effort in obfuscating
high-level types. There would be several reasons for this lack of effort.

\emph{First}, type obfuscation is unsurprisingly a side effect of data or control flow obfuscation.
Indeed, type reconstruction algorithms need both data and control flow to build type constraints,
if any of them is hidden
then the algorithms cannot work correctly. Or if the function boundary is not found
(because of anti-disassembing tricks, for example), then high-level objects cannot be recognized.
\emph{Second}, high-level types seem too coarse to be worthy
of being protected, in many cases just knowing certain values of the input which make the program
exploitable is enough. But knowledge about types expands attack surfaces because more analysis
can be proceeded, beside decompilation see the survey~\cite{caballero_type_2016} for a
more complete list.
% We exploit the \emph{semantics gap}

\section{Type flattening compiler}
We present \emph{uCc}, an open source C compiler that explicitly obfuscates high-level types, its
objective is to make reconstructing types from machine codes harder.
It is implemented in Rust, the source code and a brief user guide are given at (cite).
Our effort is to attack the core of type reconstruction algorithms,
these are the data flow used to build type constraints, and the \emph{semantic gap} between types in the high-level language and their
representations in the low-level machine code.
Beside custom tricks, \emph{uCc} uses extensively
obfuscation transformations based on Mixed Boolean Arithmetic expressions~\cite{eyrolles_obfuscation_2017,zhou_information_2007}.
To make some code diversity, the compiler is \emph{probabilistic}: given a source
code, \emph{uCc} generates each time a different (but computationally equivalent) output machine code.

We test outputs of \emph{uCc} against some decompilers, table~\ref{description} describes briefly
tested functions and their types in source codes, table~\ref{flattening} shows results of using
$4$ decompilers to decompile binaries generated by \emph{uCc}. The test suite is available at
the repository.

\begin{table}
	\begin{center}
		\caption{Functions and original types}\label{description}
		\begin{tabular}{|l|l|}
			\hline
			Description & Function type \\
			\hline
			identity & \texttt{int id(int)} \\
			\hline
			division & \texttt{int div(short, char)} \\
			\hline
			modular & \texttt{char mod(short, char)} \\
			\hline
			increase $p$ then dereference & \texttt{int inc\_deref(int *p)} \\
			\hline
			strlen  & \texttt{int slen(char*)} \\
			\hline
			sum of array $a$ & \texttt{long sum(short *a, int n)} \\
			\hline
			$n$-th fibonaci number & \texttt{int fibo(int n)} \\
			\hline
			num. of steps until $n \rightsquigarrow 1$ (Collatz's conj.)  & \texttt{int collatz(int n)} \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

\begin{table}
	\begin{adjustwidth}{-6.21in}{-5.95in}
		\begin{center}
			% \centering
			\caption{Type flattening effect}\label{flattening}
			% \begin{adjustbox}{max width=1.5\textwidth}
			\begin{tabular}{|l|l|l|l|l|}
				\hline
				Original type & Hex-Rays & JEB & Ghidra & Snowman \\
				\hline
				\texttt{int id(int)} & \texttt{int64 id(int64)} & \texttt{ulong id(ulong)} &
				\texttt{ulong id(void)} & \texttt{int64 id(int64)} \\
				\hline
				\texttt{short div(short, char)} & \texttt{int64 div(uint64, uint64)} & \texttt{div\_t div(int, int)} &
				\texttt{div\_t div(int, int)} & \texttt{int64\_t div(int64, int64)} \\
				\hline
				\texttt{char mod(short, char)} & \texttt{int64 mod(int64, int64)} & \texttt{ulong mod(ulong, ulong)} &
				\texttt{ulong mod(void)} & \texttt{int64 mod(int64, int64)} \\
				\hline
				\texttt{int inc\_deref(int*)} & \texttt{int64 inc\_deref(int64)} & \texttt{ulong inc\_deref(ulong)} &
				\texttt{ulong inc\_deref(undefined8)} & \texttt{int64 inc\_deref(uint64)} \\
				\hline
				\texttt{int slen(char*)} & \texttt{int64 slen(int64)} & \texttt{ulong slen(ulong)} &
				\texttt{ulong slen(undefined8)} & \texttt{int64 slen(uint64)} \\
				\hline
				\texttt{long sum(short*, int)} & \texttt{int64 sum(int64, uint64)} & \texttt{ulong sum(ulong, ulong)} & \texttt{undefined8 sum(undefined8)} & sn \\
				\hline
				\texttt{int fibo(int)} & \texttt{int64 fibo(uint64)} & \texttt{ulong fibo(ulong)} &
				\texttt{ulong fibo(undefined8)} & \texttt{int64 fibo(int64)} \\
				\hline
				\texttt{int collatz(int)} & \texttt{int64 collatz(int64)} & \texttt{ulong collatz(ulong)} &
				\texttt{ulong collatz(undefined8)} & \texttt{int64 collatz(int64)} \\
				\hline
			\end{tabular}
		\end{center}
	\end{adjustwidth}
\end{table}

It may worth noting that \emph{uCc} makes no effort in making disassembling or function
boundary detection hard. Indeed, functions in tested binaries are perfectly detectable,
disassemblable and decompilable.
